---
title: "Interactive web-based simulation models with deSolve and shiny"
author: "Thomas Petzoldt"
date: "2017-06-22, 2019-02-04"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("deSolve")
library("shiny")

## The following setting uses the shiny apps are
## - taken from local directories (FALSE)
## - or started from a remote server (limno-live)
```

<span style="color:red">Under construction:</span> the embedded shiny apps don't show up yet, because Github blocks embedded iframes from non-encrypted (i.e. http instead of https) connections.

## Introduction

The following examples show how to control dynamic models implemented in the 
[**R**](https://www.r-project.org) language interactively from a web interface. We use 
a technically simple ordinary differential equation (ODE) model as an example, then we will
see how to add GUI elements and finally how we can make it faster by
using compiled code.

This tutorial was originally presented at the [useR!2017](https://user2017.brussels/) 
conference in Brussels.

The HTML version embeds the shiny apps from a remote server and needs no local 
installation. The source code of this HTML file is in Rmarkdown (.Rmd) format 
that can be run locally from within RStudio when the shiny apps locally installed.

Note: this is an interactive document. <span style="color:blue">Please
use the `Code` buttons to unfold the code of individual code cunks or
for the complete document.</span> The content is best viewed with
Chrome or Firefox.

### Ingredients needed

0. A recent installation of the [**R**](https://www.r-project.org) language
1. A dynamic **deSolve** model
2. The **shiny** package and an R script **app.R**
3. A **shiny server**
    * the built-in server of [**RStudio**](https://www.rstudio.com) (on the local machine)
    * the web service [shinyapps.io](http://www.shinyapps.io/)
    * an own Linux server with shiny-server software, see  http://shiny.rstudio.com/deploy/ how to install it


### The Brusselator

The Brusselator (Brussels Oscillator) a mathematical model of an
autocatalytic reaction of two substances A and B, two intermediate products 
X and Y and two final products C and D. The latter are continuesly removed.

1. A &rarr; X
2. B + X &rarr; Y + C
3. 2X + Y &rarr;  3X
4. X &rarr; D

* developed by Ilya Prigogine and R. Lefever 1967,68 at the Université
  Libre de Bruxelles
* important theoretical example for studying oscillations and
  bifurcation
* see [Wikipedia](https://en.wikipedia.org/wiki/Brusselator) for details

$$
\begin{aligned}
\frac{dX}{dt} &= k_1 A - k_2 B X + k_3 X^2 Y - k_4 X \\
\frac{dY}{dt} &= k_2 B X - k_3 X^2 Y
\end{aligned}
$$
<hr />

### The Brusselator in **R**

Implementation of the model in **R** is straightforward. We need the package
**deSolve**, translate the differential equations of the Brusselator
into an R function, define initial values ($y_0$), parameters and time
steps and then use ode for the numerical solution. Finally we plot the
results.

```{r, fig.height=4}
library("deSolve")

brusselator <- function(t, y, p) {
  with(as.list(c(y, p)), {
    dX <- k1*A   - k2*B*X    + k3*X^2*Y - k4*X
    dY <- k2*B*X - k3*X^2*Y
    list(c(X=dX, Y=dY))
  })
}

y0 <- c(X=1, Y=1)
parms <- c(A=1, B=3, k1=1, k2=1, k3=1, k4=1)
out <- ode(y = y0, times=seq(0, 100, .1), brusselator, parms)
plot(out)
```


### A minimalistic Shiny app

```{r, eval=FALSE}
library("shiny")

server <- function(input, output) {
  output$distPlot <- renderPlot({
    hist(rnorm(input$obs))
  })
}

ui <- fluidPage(
    sliderInput("obs", "Number of observations:", min = 10, max = 500, value = 100),
    plotOutput("distPlot")
)

shinyApp(ui = ui, server = server)
```

The second part creates the web interface. The app consists of a
server element doing the computations and plotting, and a user
interface (ui), that is responsible for handling user inputs and for
sending the outputs back to the user.  Both parts, the server and the
ui can be saved in a single file 'app.R'. It can then be run directly
from **RStudio** or placed on a shiny-enabled web server.

**Exercise:** Copy the app to Rstudio, save it to the disc (fille
  app.R) and try it out

**More about this:** https://shiny.rstudio.com


## Combine the parts

Now let's combine the shiny ui/server environment with our Brusselator
model.  We start with a very simple application that allows to change
two parameters `A` and `B` (essentially the start concentrations of
substances `A` and `B`) and then plots the time series of the model.

### A first approach

```{r results='asis', echo=FALSE}
cat('<iframe src="http://limno-live.hydro.tu-dresden.de/apps/bruss01" style="border: none; width: 75%; height: 500px" />\n')
```

The code is stunningly simple, consisting of three parts:

1. The model from above (without any modifications).
2. The server component. It consists of a single function getting
`input` and `output`.  The two parameters `A` and B are taken from the
`input` and the plot is then directed to an element
`output$brussels`of the `output`.  Note also that we use deSolve's
function `matplot.0d` instead of `plot` as before, so that both state
variables appear in a single figure.
3. The ui calls then a shiny element `fluidpage` with two
   `numericInput` calls and one call to `plotOutput`.


```{r bruss1_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("bruss01/app.R")
```

```{r bruss1_2,  eval=FALSE}
<<brusselator_1>>
```


### Improvments

Now, we can start to adapt the app to our needs: change the layout,
unlock more parameters of the model, add a second figure and so on. An
especially convenient layout is the `sidebarLayout` that separates the
control elements (`sidebarPane` left or top) and output of the app
(`mainPanel` right or bottom).

HTML tags like `h2`, `h3`, or `p` can be used to annotate the app, see
help pages for more.

```{r results='asis', echo=FALSE}
cat('<iframe src="http://limno-live.hydro.tu-dresden.de/apps/bruss02" style="border: none; width: 100%; height: 800px" />\n')
```


```{r bruss2_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("bruss02/app.R")
```

```{r bruss2_2,  eval=FALSE}
<<brusselator_2>>
```


The code above should work as intended, but there are of course
further options of improvement.

1. Improve layout and design of the app: this can be quite time
   consuming and wil not be handled here.
2. Generalize the app, so that new models can instantly be
   implemented.
3. Make the model faster.

Here, we want to focus on options 2 and 3.

### Generalization

**R** is a programming language, so why not creating the user
interface directly from information found already in the initial
values and parameters of the model?  This would allow us to save
stupid typing (and copy and paste!) and to adapt the user interface
``automagically'' if we change something in the model.

The solution is, to construct lists of lists beforehand, in our case
one for the parameters `L_parms` and one for the initial values
`L_y0`. Each element of the top level list contains all the parameters
needed by a call to `numericInput`, including some constant optional
values like the `witdh` of the input field.  The templates stored in
these lists are then expanded with `lapply` and `do.call`.  The result
is exacly the same as befors, compare:
http://limno-live.hydro.tu-dresden.de/apps/bruss03


<!---
```{r results='asis', brusselator_3}
cat('<iframe src="http://limno-live.hydro.tu-dresden.de/apps/bruss03" style="border: none; width: 1024px; height: 800px" />\n')
```
--->


```{r bruss3_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("bruss03/app.R")
```

```{r bruss3_2,  eval=FALSE}
<<brusselator_3>>
```


## Make it faster

As a second improvement of our app, we want to make our model
faster. Quick response of a server based app is crucial when models
become bigger, because our users may think that the connection is
broken otherwise.

This is fortunately possible, given that some development tools
(e.g. the Rtools on Windows) are installed.

Package **deSolve** supports models in *compiled code* provided by a
`.dll` / `.so` directly.  It allows the solver to communicate directly
with the model, and without any overhead of a back-call to **R**.  Our
experience shows, that this can then be 2 ... 200 times faster,
depending on the type of model.

Several approaches were developed, depending on model complexity and
your skills:

* write it your model ``by hand'' in Fortran or C
* use a code generator:
    * package **ccSolve** from Karline Soetaert (from Github)
    * package **cOde ** from David Kaschek for smaller models
    * package **rodeo** from David Kneis for complex models

We use **cOde** here, because it is especially convenient for small
models, while for more complex models, the table based notation of
**rodeo** can be more advantageous.

### The Brusselator in C

Package **cOde** allows to compile our (technically!) simple
Brusselator without knowledge of Fortran or C.  The only prerequisite
is, that we have the development tools (the compiler etc. ...)
installed, that the system PATH correctly set, and that write access
to the working directory is granted.

Then it is sufficient, to write the differential equations in a
character vector and then to compile it with `funC`. This creates a C
file and compiles it into a shared library (.dll or .so), which then
can be numerically solved with `odeC`.

```{r bruss4_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("bruss04/brussC_test.R")
```

```{r bruss4_2,  eval=FALSE}
<<brusselator_4_core>>
```

### Shiny app with compiled code

As a final exercise, we modify our app from before, to use the
compiled C version of the model instead of the R version.

```{r bruss4_3, cache=FALSE, echo=FALSE}
knitr::read_chunk("bruss04/app.R")
```

```{r bruss4_4,  eval=FALSE}
<<brusselator_4>>
```

The result looks identical as before, just a little bit faster.

```{r results='asis', echo=FALSE}
cat('<iframe src="http://limno-live.hydro.tu-dresden.de/apps/bruss04" style="border: none; width: 100%; height: 800px" />\n')
```


If you want to see the C code: [here is it](http://limno-live.hydro.tu-dresden.de/tutorials/brusselator.c.txt)

## Further reading

* The reference site of the **deSolve** package is http://desolve.r-forge.r-project.org
* Further examples may appear at https://github.com/tpetzoldt/dynamic-R-models

----

**Original author:** [tpetzoldt](https://github.com/tpetzoldt)

