---
title: 'deSolve: Use of Compiled Code'
author: "Karline Soetaert and Thomas Petzoldt"
date: "`r Sys.Date()`"
bibliography: desolve.bib
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("deSolve")
```


## Introduction

The implementation of differential equation models directly in **R** [@RCore2018] is ideal for the development of dynamic models, because everything can be immediately tested and the whole infrastructure of **R** can be used. However, this adds considerable overhead, especially due to the automatic checking and type conversion of **R**, so that models cannot run with unlimited performance in this mode.

Fortunately, there is a way to unlock the handbreake. Both **R** in general and package **deSolve** [@Soetaert2010c;@Soetaert2010a] in particular allow to write time-critical parts in a so-called compiled language, e.g. Fortran or C/C++. Then, all solvers of **deSolve** allow direct communication between solvers and a compiled model on the machine code level, while keeping R's facilities for data management and visualisation. Details are described in the package vignette [$\rightarrow$ compiledCode](https://CRAN.R-project.org/package=deSolve/vignettes/compiledCode.pdf) of **deSolve** while the short tutorial here serves as a quick introduction.

**General principle**

* Implement core model (and only this) in C or Fortran,
* Use data handling, storage and plotting facilities of **R**.

**R** starts the solver with the compiled model, both communicate directly without the overhead of calling back to **R** and finally the return the results tu the **R** level where it is accessible by the user.

## The Lorenz system in pure R

```{r}
library(deSolve)
library(scatterplot3d)

Lorenz <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <- a * X + Y * Z
    dY <- b * (Y - Z)
    dZ <- -X * Y + c * Y - Z
    list(c(dX, dY, dZ))
  })
}

parameters <- c(a = -8/3, b = -10, c =  28)
state <- c(X = 1, Y = 1, Z = 1)
times <- seq(0, 100, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)

plot(out)
scatterplot3d(out[,-1], type="l")
```


## The Lorenz system in C


```{r engine="c", eval=FALSE}
/* file lorenz.c */
#include <R.h>
static double parms[3];
#define a parms[0]
#define b parms[1]
#define c parms[2]

/* initializer  */
void initmod(void (* odeparms)(int *, double *)) {
    int N = 3;
    odeparms(&N, parms);
}

/* Derivatives */
void derivs (int *neq, double *t, double *y, double *ydot,
             double *yout, int *ip) {
    ydot[0] = a * y[0] + y[1] * y[2];
    ydot[1] = b * (y[1] - y[2]);
    ydot[2] = - y[0] * y[1] + c * y[1] - y[2];
} 
```


### Compile the model

* needs compilers (RTools) installed and in the systems' search path

```{r}
system("R CMD SHLIB lorenzc.c")
```


### Call the model from R

```{r}
library(deSolve)
library(scatterplot3d)
dyn.load("lorenzc.dll")

out <- ode(state, times, func = "derivs", parms = parameters,
  dllname = "lorenzc", initfunc = "initmod")

dyn.unload("lorenzc.dll")

plot(out)
scatterplot3d(out[,-1], type="l")
```

Don't forget to `dyn.unload` the model afterwards, otherwise the DLL will remain 
locked and cannot be overwritten in a next compiler run.


## The Lorenz system in Fortran


<!---

Note:

Rmarkdown is capable to call fortran and c code directly from the code chunks, 
so it is (theoretically and practically) not necessary to have the c/fortran files 
on the hard disc. It then compiles it to a temprary dll 
(with a cryptic random file name) and loads this directly.

I don't use this here, intentionally. T.P.
--->


```{r, engine="fortran", eval=FALSE}
c file lorenz.f
       subroutine initmod(odeparms)
         external odeparms
         double precision parms(3)
         common /myparms/parms

         call odeparms(3, parms)
         return
       end

       subroutine derivs (neq, t, y, ydot, yout, ip)
         double precision t, y, ydot, a, b, c
         integer neq, ip(*)
         dimension y(3), ydot(3), yout(*)
         common /myparms/a,b,c

         ydot(1) = a * y(1) + y(2) * y(3)
         ydot(2) = b * (y(2) - y(3))
         ydot(3) = -y(1) * y(2) + c * y(2) - y(3)

        return
       end
```


### Compile the model

* needs compilers (Rtools) installed and in the systems' search path
* this is quite standard on Linux, whereas Windows users have to download the [Rtools](https://cran.r-project.org/bin/windows/Rtools/) separately.

```{r}
system("R CMD SHLIB lorenzf.f")
```


### Call the model from R

```{r}
dyn.load("lorenzf.dll")

out <- ode(state, times, func = "derivs", parms = parameters,
  dllname = "lorenzf", initfunc = "initmod")

dyn.unload("lorenzf.dll")

plot(out)
scatterplot3d(out[,-1], type="l")
```

Don't forget to unload the model afterwards, otherwise the DLL will remain 
locked and cannot be overwritten in a next compiler run.

## Benchmark

Run the 3 versions (R, C, Fortran) several times in an R script and make a benchmark.

```{r eval=FALSE}
library(deSolve)
library(scatterplot3d)

## put the pure R model here

## compile the C code
system("R CMD SHLIB lorenzc.c")

## compile the fortran code
system(..............)

## load the DLLs (note .so on Linux)
dyn.load("lorenzc.dll")
dyn.load(............)

## run the R model 10 (or 100) times
system.time(
  for(i in 1:10) out   <- ode(state, times, Lorenz, parms = parameters)
)

## run the C model 10 (or 100) times
..............

## run the Fortran model 10 (or 100) times
..............

## don't forget to unload the DLLs (.so on Linux)
dyn.unload("lorenzc.dll")
dyn.unload("lorenzf.dll")
```

## Outlook

More can be found in the package vignette [compiledCode](https://CRAN.R-project.org/package=deSolve/vignettes/compiledCode.pdf) or in the paper of @Soetaert2010a. In addition to this, compiled differential equation models are also supported by other packages like **cOde** and **dMod** [@Kaschek2018;@Kaschek2018b] or **rodeo**  [@Kneis2017;@rodeo2018].



## References

<div id="refs"></div>

----

**Copyright and original author:** [tpetzoldt](https://github.com/tpetzoldt), `r Sys.Date()`

