---
title: "Efficient ODE Models with rodeo"
author: "Thomas Petzoldt"
date: "2017-06-30, 2019-02-06"
bibliography: rodeo.bib
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    #code_folding: hide
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library("deSolve") 
library("shiny") 
``` 

Package **rodeo** [@Kneis2017;@rodeo2018]
combines a series of complementary approaches to implement differential equation models 
(currently ODEs and 1D PDEs) in an efficient way:

1. The stoichiometry matrix notation ([Petersen matrix](https://en.wikipedia.org/wiki/Petersen_matrix))
2. A code generator to generate **R** or **Fortran** code automatically
3. Overhead-free access to forcing data from Fortran code,
4. Syntax for one-dimensional models, built-in integrator, documentation support, ...


A detailed description and tutorial can be found in the comprehensive [**rodeo package
vignette**](https://CRAN.R-project.org/package=rodeo/vignettes/rodeoVignette.pdf)
from [David Kneis](https://github.com/dkneis). The following example is served as a small teaser.

## A Predator-prey example

As a first example, we use a predator-prey model with 
three differential equations of the following form:


#### Resource (Substrate)
$$\frac{dS}{dt} = s_{in}(t) - b \cdot S \cdot P + g \cdot K$$

#### Prey (Producer)
$$\frac{dP}{dt} = c \cdot S \cdot P - d \cdot K \cdot P$$  

#### Predator (Consumer)
$$\frac{dK}{dt} = e \cdot K \cdot P - f \cdot K$$


This is intentionally very basic, but we can see already, that the
equations are coupled. As an example, the term $K \cdot P$, i.e. the
interaction between prey and predator, appears twice, as a loss in the
prey equation and as a gain in the predator equation. Only the rate
constants $d$ and $e$ are different. A similar connection exists
between resource and prey.

The obvious redundancy of terms (that can be much more complex for
bigger models) can then be used to split the system of equations in
two parts, a process vector and a stoichiometry matrix.

These two parts are essentially tables and can be organised as
.csv-format or even in Excel (or LibreOffice) worksheets.

## Tabular model definition

In the following, we load the required packages, **readxl** for
reading the tables, **deSolve** for the numerical solution of the
equations and **rodeo** that creates an **R** [@RCore2018] or **Fortran** code from
this.  Package **rodeo** is implemented in the **R6** class system of
**R**, we can see this from the `new` constructor and the dollar notation of the
methods.

```{r}
library("readxl")
library("deSolve")
library("rodeo")
```

We load now the Excel tables to **R** and show its contents, but it is
of course also a good idea to open the [`model.xlsx`](model.xlsx)-file directly.

A typical **rodeo** model needs the following tables:

* `vars` for the state variables, 
* `pars` for the model parameters, 
* `funs`for some helper functions and 
* `stoi` for the stoichiometry matrix. 

Tables `xstoi` and `forc` are optional and will be explained later. The last two 
lines in the following code convert the list of tables in the appropriate format. They are not
necessary if we import the tables from .csv instead of xlsx.

```{r eval=TRUE, echo=TRUE}
sheets <- c("vars", "pars", "funs", "pros", "stoi", "xstoi", "forc")

## read the tables
tables <- lapply(sheets, function(sheet) read_excel("model.xlsx", sheet = sheet))

## convert from read_xl specific "tibble" class to data.frame
tables <- lapply(tables, as.data.frame)
names(tables) <- sheets
```

## Now, let's inspect the tables

```{r eval=TRUE, echo=TRUE}
tables$vars
```


```{r eval=TRUE, echo=TRUE}
tables$pars
```


```{r eval=TRUE, echo=TRUE}
tables$pros
```

The stoichiometry table can be given either datbase-like or as a cross
table in matrix form.  Both notations have their pros and cons. While
database tables (`stoi` format) contain less redundancy and are easier
to annotate, the interactions between state variables and processes
are easier to follow in the crosstabular notation.

```{r eval=TRUE, echo=TRUE}
tables$xstoi
```

## Code generation and compile

Now we create the **rodeo** object by calling the constructor `new`
with the given data frames from the `tables` list. The argument `dim=1`
states that it is a box-model in 1 dimension, i.e. it is essentially
zero dimensional.

```{r, eval=TRUE, echo=TRUE}
## create rodeo object
model <- with(tables, 
  rodeo$new(vars=vars, pars=pars, funs=funs, pros=pros, stoi=stoi, dim=1))
```

The same can also be done from the cross-tabular form (argument:
`asMatrix`). We only need to make sure that the column names contain
all state variables and the row names all corresponding processes.


```{r eval=FALSE}
xxstoi <- as.matrix(tables$xstoi[,-1])
rownames(xxstoi) <- tables$xstoi[,1]
model <- with(tables, 
  rodeo$new(vars=vars, pars=pars, funs=funs, pros=pros, stoi=xxstoi, 
            asMatrix=TRUE, dim=1)
)
```

As said, the example shown here is intentionally basic. However, it shows 
from the very beginning how to use external forcings, i.e. data read from an 
external file (see [input.txt](input.txt)).

```{r eval=TRUE, echo=TRUE, collapse=TRUE}
tables$forc
```

Here `name` is the name of the function that can be used in the process
vector, column is the column name in the file, mode is the
interpolation mode. A negative value means linear interpolation.

If such functions (and of course also corresponding files) exist, an
appropriate Fortran module can be generated with `forcingFunctions`:

```{r}
fforc <- forcingFunctions(tables$forc)
write(fforc, file = "forcings.f95")
```

Now we can compile the model, given that the R developer tools
(**Rtools**) are installed.  A second source file `functions.f95`
contains additional formal definitions and (optionally) user-provided
functions written directly in Fortran. It is therefore user-editable,
while other parts (forcings and some includes taken directly from the
package) should be left untouched.

```{r}
ret <- model$compile(sources  = c("forcings.f95", "functions.f95"))
```

The result of the compilation process is now found in a temporary
folder, depending on the type of the operating system. It can be
located with `tempfile()`, and it is of course also possible to
archive the compiled `.dll/.so` for later use, but that's another
topic and is not covered here.

## Model simulation

Now we are ready to assign parameters (`setPars`) and initial values
(`setVars`) to run the model. Both, `vars` and `pars`must be **named**
numeric vectors. They can be either taken from optional columns of the
Excel tables or assigned separately (see 2nd scenario).

```{r}
vars <- setNames(tables$vars$default, tables$vars$name)
pars <- setNames(tables$pars$default, tables$pars$name)

model$setPars(pars)
model$setVars(vars)
```

The numerical solution is then carried out with function `dynamics` that calls an
appropriate numerical solver from the **deSolve** [@Soetaert2010a;@Soetaert2010c] package. 
Now, let's run some scenarios, a first scenario with the default
parameter set:


```{r, fig.height=3}
out1 <- model$dynamics(seq(0, 200, 0.1))
plot(out1, which=c("S", "P", "K"), mfrow=c(1,3))
```

... a second scenario with a user-supplied parameter vector:

```{r, fig.height=3}
pars <- c(S_in_def=0.1, b=0.1, c=0.1, d=0.1, e=0.1, f=0.1, g=0.0)
model$setPars(pars)
out2 <- model$dynamics(seq(0, 200, 0.1), method="bdf")
plot(out2, which=c("S", "P", "K"), mfrow=c(1, 3))
```

... and finally a third scenario where the forcing data file
[`input.txt`](input.txt) is enabled by setting the default value (`input_default`)
to `NaN`. A model with external forcing is called a non-autonomous
model. In such cases it is always a good idea to limit the step size
by setting `hmax` to a value equal or smaller than the temporal
resolution of the input data.

```{r, fig.height=3}
pars <- c(S_in_def=NaN, b=0.1, c=0.1, d=0.1, e=0.1, f=0.1, g=0.0)
model$setPars(pars)
out3 <- model$dynamics(seq(0, 200, 0.1), method="bdf_d", hmax=1)
plot(out1, out2, out3, which=c("S", "P", "K"), mfrow=c(1,3))
```

## Controlling numerical precision

The precistion of **deSolve** solutions is controlled with two optional arguments 
`atol` and `rtol`. This is particular importance if scales of states differ, e.g. 
if a resource is in the range of 0.0001 while the bacteria may have an abundance of $10^9$.

As the table notation allows additional user-defined columns, individual 
state-variable dependent tolerances can be easily set.

```{r}
pars <- c(S_in_def=NaN, b=0.1, c=0.1, d=0.1, e=0.1, f=0.1, g=0.0)
model$setPars(pars)
out4 <- model$dynamics(seq(0, 200, 0.1), method="bdf_d", hmax=1, 
                       atol=tables$vars$atol, rtol=tables$vars$rtol)
plot(out4)
```



## Documentation support

The package contains several convenience functions, e.g. to visualize
the stoichiometry matrix graphically or to export the parameter,
variable, process and stoichiometry tables to Latex or HTML for
documentation and publication.

```{r}
model$plotStoichiometry(box=1)
```



```{r}
cat(exportDF(tables$pros[c("name", "unit", "expression", "tex")], tex=TRUE,
         funCell=c(name=function(x){paste0("\\textit{",x,"}")},
           tex=function(x){paste0("$",x,"$")})))
```


## Further reading

O.k., this was a first round. More technical background can be found in @Kneis2017, in the package the help files and
in the extensive [rodeo package
vignette](https://CRAN.R-project.org/package=rodeo/vignettes/rodeoVignette.pdf).
And last but not least a web-based shiny examples, e.g. for a  [shallow lake](http://limno-live.hydro.tu-dresden.de/models/shallowlake/) and [WebLAB](http://limno-live.hydro.tu-dresden.de/weblab/) demonstrate some more complex rodeo-models.


## References

<div id="refs"></div>

----

**Copyright and original author:** [tpetzoldt](https://github.com/tpetzoldt), `r Sys.Date()`


