---
title: "An individual-based model of a flow-through system"
author: "Thomas Petzoldt"
date: "2019-02-04"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(simecol)
library(plyr)
library(ggplot2)
```

## Introduction

Note: this is an interactive document. <span style="color:blue">Please
use the `Hide` resp. `Code` buttons to fold or unfold the code of individual code cunks or
of the complete document.</span> The content is best viewed with
Chrome or Firefox.

## General approach

Let's assume the following system of ordinary differential equations for a 
flow-through bioreactor (chemostat), where $X$ is the concentration of organisms (e.g. cells/volume)
in the system growing on a limited resource $S$ (e.g. nutrient/volume). The instantanaeous growth rate $r$ (1/time) depends 
on a maximum growth rate $r_{max}$ and the resource concentration $S$ via a Monod 
equation, with half-saturatio constant $k_s$.
Fresh medium with resource concentration $S_0$ is imported with a dilution rate 
$D$ that controls also export of cells and of remaining medium. A stoichiometric 
coefficient ($Y$, yield) converts between concentration units of cells and nutrients:

$$
\frac{dX}{dt} = (r - D) \cdot X\\
\frac{dS}{dt} = D \cdot (S_0-S) - \frac{S}{Y}\\
\text{with:}\\
r = r_{max} \cdot \frac{S}{k_s+S}
$$
The above system of equations applies for a continuous system, but there may be 
reasons to adapt (or approximate) such a model to a discrete-time step and a 
discrete number of individuals. One reason may be a relatively small number of organisms 
(e.g. a few hundred rotifers instead of 10^8 cells of bacteria) or the nesseccity to describe
individual variation (e.g. age, size, cell content) of the organisms.

This can be done with a so-called *individual-based model*, where the equations of such a system are, in principle, quite similar. 
Its main differences are:

* the population is represented by a countable number ($N$) of individuals instead of a state variable $X$,
* the model is solved as a stochastic process where probabilities are used instead of rate parameters (1/time),
* an event-like fixed time step scheme is used instead of a continuous integrator.


## Plain implementation

The following example implements a very basic individual-based approach. The population of 
individuals is implemented as a data frame (`inds`) with, in this example, only one single column (`age`).
Substrate $S$ is implemented as a non-individual-based pool variable, the parameters are in a vector (`parms`).

All life-actvities are carried out in one `live`-function. Mortality and cell division are realized by
subsetting randomly selected surviving individuals from the population, or by 
adding copies of randomly selected "dividing cells" to the population. 
The `age` of divided cells is set to zero for both daughters of each parent cell.

Mortality and substrate input depend on $D$, while cell division depends on the 
substrate concentration. A safeguard check prevents $S$ for becoming negative and
a control parameter `DELTAT` allows an adaption of the time step.

The solution of the model is then straightforward by iterating the `live`-function in a for-loop.

```{r chem1_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("chemostat-ibm-plain.R")
```

```{r chem1_2,  eval=TRUE}
<<chemostat_1>>
```

The simulation shows then a typical time series. The curves show a stochastic 
pattern, that can be reduced if the sample size (the number of individuals) is 
increased to, e.g. 10.000 or more. The yield is set to $Y=1$ to make comparisons easier.


```{r chem1_3}
matplot(o$time, o[c("N", "S", "maxage")], xlab="time", ylab="N, S", type="l")
head(o)
```

## Object orientation with **simecol**

The above implementation is extremely simple and may already serve as a starting 
point for extensions. However, almost all variables are held in global work space,
which can become cumbersome and confusing when the application grows.

Package **simecol** @Petzoldt2007 suggests an object oriented programming (OOP) 
approach for such cases. In contrast to other OOP approaches for IBMs that model
individuals as objects in their habitat, in an ecosystem and so forth, simecol "models
the components of a simulation model". This is quite pragmatic and aims to make 
comparison of different simulations (scenarios, instances) easier. Details of the 
approach are found in @Petzoldt2007. 

In the example below, all different processes of the live-loop are subdivided in 
separated functions in the `equations`-slot of the object that are then called 
from the `main` function. Parameters (`parms`), the `init`-values of the state variables 
(a list with `inds` the table of individuals and `S` the substrate pool),
and the `times` are encapsulated in the object. The solver is set to `iteration` 
that is built-in in the package.


```{r chem_2_1, cache=FALSE, echo=FALSE}
knitr::read_chunk("chemostat-ibm-simecol.R")
```

```{r chem_2_2, eval=TRUE}
<<chemostat_2>>
```

### Simulation

The simulation of the model can then, in principle, be triggered by:

```{r chem2_2a, eval=FALSE}
ibm_test <- sim(ibm_test)
o <- out(ibm_test)
```

### Observer function

In this case, the complete set of state variables (i.e. S and the complete data 
frame of inds) ist stored for each time step, so that ibm_test contains both,
the original model (with equation and parameters) and all simulation results, 
that can be extracted with out(). It is then up to the user to analyze the results from out.

The analysis of results can already be made during the simulation by providing an 
optional `observer`-function. It gets the full state information and aggregates the
required parts, and it can also be used for printing status information or for
animated graphics during the simulation.

```{r chem_2_3, eval=TRUE}
<<chemostat_3>>
```

The simulation can then be run  with the `sim`-function:


```{r chem_2_4, eval=TRUE}
ibm_test <- sim(ibm_test)
o <- out(ibm_test)
matplot(o$time, o[c("N", "S", "age")], xlab="time", ylab="N, S", type="l")
```

### Repeated simulations

As the IBM is a stochastic model, it would make sense to repeat it seveal times. 
The following gigure shows four independent simulation runs:

```{r chem_2_5, eval=TRUE}
par(mfrow=c(2,2), mar=c(4,3,1,1)+0.1)
ret <- lapply(1:4, FUN=function(x) cbind(run=x, out(sim(ibm_test))))
ret <- lapply(ret, function(o) matplot(o$time, o[c("N", "S")], xlab="time", ylab="N, S", type="l"))
```

... and the next one a binned representation of 100 simulations. 

```{r}
library(plyr)
library(ggplot2)
ret <- lapply(1:100, FUN=function(x) cbind(run=x, out(sim(ibm_test))))
longtab <- rbind.fill(ret)

ggplot(longtab, aes(time, N)) + geom_bin2d(binwidth=c(1,10))
ggplot(longtab, aes(time, S)) + geom_bin2d(binwidth=c(1,10))
```

Here, `lapply`returns a list of 100 data frames with the run number in the first column.
Function `rbind.fill` (from package **plyr**) then concatenates all these data 
frames together into tidy database format `longtab` that can then be plotted with
one of the ggplot functions.

### Equilibrium






